Verifying Concurrent Search Structure Templates for C Programs using VST

### 2023-04-14 concurrency/CONCUR 2023/paper.tex
initial writing proof of give-up traverse

\newcommand\dboxed[1]{\dbox{\ensuremath{#1}}}
\newcommand{\master}[2]{\ensuremath{\mathrm{Master}_{#1}(#2)}}
\newcommand{\snap}[1]{\ensuremath{\mathrm{Snapshot}(#1)}}
\newcommand{\ghost}[2]{\ensuremath{\dboxed{#1}^{#2}}}
\newcommand{\us}{$\mu$s}
\newcommand{\gnamety}{\ensuremath{\mathsf{gname}}}
\newcommand{\treerep}{\ensuremath{\mathsf{Node}}}
\newcommand{\nodeboxrep}{\ensuremath{\mathsf{Node\_ref}}}
\newcommand{\lockinv}{\ensuremath{\mathsf{lock\_inv}}}

@@

The specification states that if the node \texttt{n} is within the footprint of the search structure, then the \texttt{traverse} function operates on the entire search structure and returns a subsequent node \texttt{n'}, where \texttt{k} falls within the range of node \texttt{n'}.

The key to the verification of \texttt{traverse} function is the loop invariant for the top-level loop. We define the loop invariant for \texttt{traverse} function of lock-coupling style as follows: 
\begin{align*} &\texttt{traverse\_inv(pn)} \triangleq\ \exists \texttt{n'}.\ (\texttt{k} \in \texttt{range})\land \nodeboxrep(\texttt{pn->n'})  \ast  \texttt{R(pn->n')}   
\end{align*}

Since the \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL}. Alternatively, it terminates if \lstinline{findNext} fails to find a next node that includes the key value $\texttt{k}$ in the set of key values. Therefore, we also define loop invariants for such these cases as follows: 
\begin{align*} 
	\texttt{traverse\_inv\_1(pn)} \triangleq\ \exists \texttt{n'}.\ (\texttt{k} \in \texttt{range})\land \nodeboxrep(\texttt{pn->n'})  \ast \texttt{R(pn->n')} \ast (\texttt{pn->n'->t} = \texttt{NULL})
	\\
	\texttt{traverse\_inv\_2(pn)} \triangleq\  \exists \texttt{n'}.\ (\texttt{k} \in \texttt{range})\land \nodeboxrep(\texttt{pn->n'})  \ast \texttt{R(pn->n')} \ast (\texttt{pn->n'->t} \neq \texttt{NULL})
\end{align*}

More specifically, $\texttt{traverse\_inv\_1 (pn)}$ is true when the \lstinline{traverse} function successfully locates the subsequent node during traversal, as indicated by the condition $\texttt{pn->n'->t} = \texttt{NULL}$. In the context of a linked-list data structure, a \lstinline{t} field with a value of \lstinline{NULL} signals that the \lstinline{traverse} function has arrived at the list's final node.

In contrast, $\texttt{traverse\_inv\_2 (pn)}$ is true when the \lstinline{traverse} function fails to locate the next node during traversal, which means $\texttt{pn->n'->t} \neq \texttt{NULL}$. For instance, when trying to insert a new node with a specific key \lstinline{k} into a linked list, if the \lstinline{traverse} function cannot identify the next node, it implies that the current node has a target key equal to its value. Since the node holds the key, the \lstinline{t} field should not be \lstinline{NULL} .

@@

$$\forall \texttt{t}.\ \left\langle \nodeboxrep \texttt{(pn)} \ast \lockinv \texttt{(pn->p->lock,R'}) \ast \texttt{R(pn->p)} \ | \ \treerep\ \texttt{t} \right\rangle $$
$$\texttt{traverse(pn, k)}$$
$$\left\langle \texttt{n'}.\ \nodeboxrep \texttt{(pn)} \ast \lockinv \texttt{(pn->n'->lock,R')} \ast \texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range})
\ |\ \treerep\ \texttt{t} \right\rangle$$

\section{Proof Mechanization}
We have veried the give-up and lock-coupling template algorithms.
We describe the mechanized formal verification of the correctness of such these templates, implemented in C and verified with the VST, using concurrent separation
logic (CSL) to reason about the behavior of the C code, and integrated Iris-style logical atomicity into VST. All proofs are formalized in Coq using VST, and can be found online at ...

>>>>

\newcommand\dboxed[1]{\dbox{\ensuremath{#1}}}
\newcommand{\master}[2]{\ensuremath{\mathrm{Master}_{#1}(#2)}}
\newcommand{\snap}[1]{\ensuremath{\mathrm{Snapshot}(#1)}}
\newcommand{\ghost}[2]{\ensuremath{\dboxed{#1}^{#2}}}
\newcommand{\us}{$\mu$s}
\newcommand{\gnamety}{\ensuremath{\mathsf{gname}}}
\newcommand{\treerep}{\ensuremath{\mathsf{Node}}}
\newcommand{\nodeboxrep}{\ensuremath{\mathsf{Node\_ref}}}
\newcommand{\lockinv}{\ensuremath{\mathsf{lock\_inv}}}
\newcommand{\infp}{\ensuremath{\mathsf{InFP}}}

@@

%The specification states that if the node \texttt{n} is within the footprint of the search structure, then the \texttt{traverse} function operates on the entire search structure and returns a subsequent node \texttt{n'}, where \texttt{k} falls within the range of node \texttt{n'}.

The crux of verifying the \texttt{traverse} function lies in the loop invariant for the top-level loop. We define the loop invariant for the lock-coupling style \texttt{traverse} function as follows:
\begin{align*} &\texttt{traverse\_inv(pn)} \triangleq\ \exists \texttt{n'}.\ (\texttt{k} \in \texttt{range})\land \nodeboxrep(\texttt{pn->n'})  \ast  \texttt{R(pn->n')}   
\end{align*}
The \lstinline{traverse} function stops when it reaches the base case, which happens when the current node is \lstinline{NULL}. Alternatively, the function terminates if \lstinline{findNext} fails to identify a subsequent node with the key value $\texttt{k}$ within its set of key values. In addition to the \texttt{traverse\_inv} we defined earlier, we also need to establish loop invariants for these particular situations as follows:
\begin{align*} 
	\texttt{traverse\_inv\_1(pn)} \triangleq\ \exists \texttt{n'}.\ (\texttt{k} \in \texttt{range})\land \nodeboxrep(\texttt{pn->n'})  \ast \texttt{R(pn->n')} \ast (\texttt{pn->n'->t} = \texttt{NULL})
	\\
	\texttt{traverse\_inv\_2(pn)} \triangleq\  \exists \texttt{n'}.\ (\texttt{k} \in \texttt{range})\land \nodeboxrep(\texttt{pn->n'})  \ast \texttt{R(pn->n')} \ast (\texttt{pn->n'->t} \neq \texttt{NULL})
\end{align*}

More specifically, $\texttt{traverse\_inv\_1}$ is true when the \lstinline{traverse} function successfully locates the subsequent node during traversal, as indicated by the condition $\texttt{pn->n'->t} = \texttt{NULL}$. In the context of a linked-list data structure, a \lstinline{t} field with a value of \lstinline{NULL} signals that the \lstinline{traverse} function has arrived at the list's final node. In contrast, $\texttt{traverse\_inv\_2}$ is true when the \lstinline{traverse} function fails to locate the next node during traversal, which means $\texttt{pn->n'->t} \neq \texttt{NULL}$. For instance, when trying to insert a new node with a specific key \lstinline{k} into a linked list, if the \lstinline{traverse} function cannot identify the next node, it implies that the current node has a target key equal to its value. Since the node holds the key, the \lstinline{t} field should not be \lstinline{NULL} .

@@

$$\forall \texttt{t}.\ \left\langle  \infp (\texttt{pn->n}) \ | \ \treerep\ \texttt{t} \right\rangle$$
$$\texttt{traverse(pn, k)}$$
$$\left\langle \texttt{res} \ \texttt{n'}. \ (\texttt{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL})) \ast  \infp (\texttt{pn->n'}) \ast \texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range})
\ |\ \treerep\ \texttt{t} \right\rangle$$

Compared to the more complex specification of the lock-coupling style \texttt{traverse} function, the specification of the give-up template \texttt{traverse} function is simpler. The local pre-condition for the specification only includes $\infp (\texttt{pn->n})$, a footprint predicate indicating that node \texttt{n} is within the search structure's footprint. The \texttt{traverse} function then operates on the entire search structure and returns a subsequent node \texttt{n'}, where \texttt{k} falls within the range of node \texttt{n'}. Consequently, the local post-condition is expected to assert that \texttt{n'} is also situated within the footprint predicate and that the resource \texttt{R(pn->n')} contains the concrete information about node \texttt{n'}.

\section{Proof Mechanization}
We have veried the give-up and lock-coupling template algorithms.
We describe the mechanized formal verification of the correctness of such these templates, implemented in C and verified with the VST, using concurrent separation
logic (CSL) to reason about the behavior of the C code, and integrated Iris-style logical atomicity into VST. All proofs are formalized in Coq using VST, and can be found online at

\href{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}{\color{blue}{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}}
